SDSS-IV MaNGA: The Radial Profile of Enhanced Star Formation in Close Galaxy Pairs

### 2020-10-22 maintex.tex

% AASTeX v6.2
%\documentclass[preprint]{aastex62}

@@

Indeed, \citet{Barrera-Ballesteros:2015} used the CALIFA (Calar Alto Legacy Integral Field Area) survey to study a sample of 103 paired galaxies by varying the size of the aperture through which the \ewha\ is extracted from. In this study, they found a moderate enhancement to the sSFR in the centers of paired galaxies and a moderate suppression to the sSFR in outskirts of the paired galaxies. 
%
\citet{Pan:2019} used the SDSS-IV MaNGA survey to study radial profiles of a sample of 205 paired galaxies. The enhancement to the sSFR was shown to be the strongest in the centers of the paired galaxies. This central enhancement linearly fell with increasing galactocentric radii; however, a moderate enhancement to the sSFR remain in the outskirts of the galaxies. \citet{Pan:2019} further studied the paired galaxies as a function of merger stage; from well separated pairs to post-merger galaxies. Across the different merger stages, the sSFR enhancement was greatest in close pairs with tidal features and in post-merger galaxies. This was in agreement with previous hydrodynamical simulations which showed that a burst of star formation is triggered after the first pericenter and as the two galaxies begin to coalesce \citep{Scudder:2012}. 
%
The radial profile of sSFR in post-merger galaxies has also been studied with the MaNGA survey by \citet{Thorp:2019}. The post-merger galaxies were shown to have a strong enhancement to the sSFR in their centers as well as a moderate enhancement in their outskirts.

@@

Among 6142 MaNGA IFUs considered here, 404 have multiple galaxy objects within the IFU. We further restrict the sample by setting a relative velocity cut of $\Delta v$ $<$ 500 km s$^{-1}$ to remove projected companions. Given redshift range of the MaNGA sample and the size of MaNGA's IFUs, the maximum projected separation for a companion galaxy in the IFU is $\sim$40 kpc. Again, the galaxy sample is also restricted to galaxies with a stellar mass range of \logm\ $=$ 9.0$-$11.5 and galaxies in the Primary or Secondary MaNGA subsamples. We also require that the galaxies are classified as star forming as described in Section \ref{sec:data}. These requirements reduce our sample down to 87 star forming pairs.

Several of the target galaxies in our sample have multiple companions. From the whole set of 404 IFUs, there are 327 pairs, 67 triplets, 7 quadruplets, and 1 quintuplet. In cases where a MaNGA target has multiple viable companions galaxies, we chose the closest companion when measuring the mass ratio and projected separation.

@@

@@

@@

%%% SPFIT
For each MaNGA 1\arcsec$\times$1\arcsec\ spaxel, we use our spectral fitting code, {\sc spfit}, to get measurements of emission line fluxes, equivalent widths, velocities, and velocity dispersions of 19 emission lines and measurements of stellar masses, ages, [Fe/H], and the kinematics of the stellar populations.

%%% Reddening
We correct the emission lines for reddening using the extinction curve from \citet{Cardelli:1989} with updated coefficients from \citet{ODonnell:1994}. The extinction is parameterized as $R_V$ $\equiv$ $A_V/E(B-V)$ $=$ 3.1, where we estimate the value of the $V$-band extinction, $A_V$, by comparing the H$\alpha$/H$\beta$ ratio to the expected value of 2.85 for case-B recombination.

@@

@@

%We use our catalog of photometric objects and the \citet{Simard:2011} catalog to mask foreground/background objects within the MaNGA IFUs which may contaminate the data. Objects in our photometric catalog are masked out with a 2\arcsec\ circular aperture and the galaxies from \citet{Simard:2011} are masked out with a 2.0 R/\reff\ elliptical aperture, using the r-band S\'ersic apertures from the catalog. This gives us an individual radial profile of the sSFR for all paired and control galaxies.

@@

In the first method, paired and control galaxies are grouped into five evenly spaced stellar mass bins between \logm\ $=$ 9.0$-$11.5. Within each stellar mass bin, we create a median profile of log(sSFR) as a function of \reff-scaled galactocentric distance for both paired and control galaxies. The control profiles are constructed using all available control galaxies within each mass bin. The error associated with the ``stacked'' profile is the standard error of mean of the data at each radius bin. We show these stacked profiles for control galaxies in the left hand panel of Figure \ref{fig:ssfr_prof} and the stacked profiles for paired galaxies in the middle panel of Figure \ref{fig:ssfr_prof}. {\bf QUESTION: for the inside-IFU pairs, did you use only the primaries to build these stacked profiles? If so, this needs to made clear in the text.} Star formation in the control galaxies are quenched in their centers with respect to their disks and the difference between the sSFR in their centers and their disks increases with stellar mass, consistent with previously published results using the MaNGA survey \citep{Belfiore:2018}. In contrast, the paired galaxies show flatter sSFR profiles where the level of the sSFR remains roughly consistent across their disks except for paired galaxies with stellar masses above \logm\ $=$ 10.5, which still show some central quenching.

@@

Alternatively, the central enhancement of the sSFR in the paired galaxies can be interpreted as the quenching of star formation in the centers of isolated galaxies. While the mechanism for quenching in galaxies remains under debate, the merger process seems to be capable of rejuvenating star formation in the centers of these galaxies, resulting in the flatter sSFR profiles. 
%Further, since the sSFR enhancement in the difference profiles is largely driven by the quenched controls, it may be more appropriate to describe isolated galaxies as being centrally quenched in comparison to paired galaxies or that the profiles of paired galaxies are flattened in their centers with respect to their controls.

@@

%To emphasize the difference between primary and secondary companions and to further investigate the sSFR enhancement seen in the outskirts of low mass galaxies (\logm\ $=$ 9.0$-$9.5), we plot the difference profiles split both by stellar mass and mass ratio in Figure \ref{fig:ssfr_massdm}. The difference profile of primary companions is centrally enhanced by 0.25 dex and remains enhanced across their disks with an enhancement of 0.1-0.3 dex. The difference profile of secondary companions is centrally enhanced by 0.3 dex and is suppressed by 0.1 dex in their outskirts. Specifically looking at the paired galaxies in the mass range, \logm\ $=$ 9.0$-$9.5, the primary companions in the mass range show strong enhancements to their outskirts while the primary companions in the mass range show a weak suppression to their outskirts.

@@

%Among the three parameters, stellar mass, mass ratio, and projected separation, the projected separation has the strongest effect on the $\Delta$log(sSFR) enhancement, followed by the mass ratio. From this we see that the paired galaxies with the strongest enhancements to the central sSFR are paired galaxies with 1:1 mass ratios and  which are within 10 kpc.

%Between the splitting the $\Delta$log(sSFR) profile by stellar mass, mass ratio, and projected separation, we see that the general shape of the profile within the inner 1.0 \reff\ is unchanged. The profile is peaked in the center of the paired galaxies and linearly (in log space) to zero around 1.0 \reff. Different projected separations and mass ratios may elevate or suppress the level of the enhancement in the profile's center; however, the shape of the profile is largely unchanged.

>>>>

@@

Indeed, \citet{Barrera-Ballesteros:2015} used the CALIFA (Calar Alto Legacy Integral Field Area) survey to study a sample of 103 paired galaxies by varying the size of the aperture through which the \ewha\ is extracted from. In this study, they found a moderate enhancement to the sSFR in the centers of paired galaxies and a moderate suppression to the sSFR in outskirts of the paired galaxies.

\citet{Pan:2019} used the SDSS-IV MaNGA survey to study radial profiles of a sample of 205 paired galaxies. The enhancement to the sSFR was shown to be the strongest in the centers of the paired galaxies. This central enhancement linearly fell with increasing galactocentric radii; however, a moderate enhancement to the sSFR remain in the outskirts of the galaxies. \citet{Pan:2019} further studied the paired galaxies as a function of merger stage; from well separated pairs to post-merger galaxies. Across the different merger stages, the sSFR enhancement was greatest in close pairs with tidal features and in post-merger galaxies. This was in agreement with previous hydrodynamical simulations which showed that a burst of star formation is triggered after the first pericenter and as the two galaxies begin to coalesce \citep{Scudder:2012}.

The radial profile of sSFR in post-merger galaxies has also been studied with the MaNGA survey by \citet{Thorp:2019}. The post-merger galaxies were shown to have a strong enhancement to the sSFR in their centers as well as a moderate enhancement in their outskirts.

@@

Among 6142 MaNGA IFUs considered here, 404 have multiple galaxy objects within the IFU. We further restrict the sample by setting a relative velocity cut of $\Delta v$ $<$ 500 km s$^{-1}$ to remove projected companions. Given redshift range of the MaNGA sample and the size of MaNGA's IFUs, the maximum projected separation for a companion galaxy in the IFU is $\sim$40 kpc. Again, the galaxy sample is also restricted to galaxies with a stellar mass range of \logm\ $=$ 9.0$-$11.5 and galaxies in the Primary or Secondary MaNGA subsamples. We also require that the galaxies are classified as star forming as described in Section \ref{sec:data}. These requirements reduce our sample down to 87 star forming MaNGA targets with inside-IFU companions. While the companions to the inside-IFU MaNGA targets contained are within the MaNGA IFUs, we only use the MaNGA targets when building radial profiles. We leave out the identified companions from the analysis because the MaNGA targets are covered by the NSA catalog with geometries and total stellar masses while the companions are not covered by the NSA catalog.

Several of the target galaxies in our sample have multiple companions. From the whole set of 404 IFUs, there are 327 pairs, 67 triplets, 7 quadruplets, and 1 quintuplet. In cases where a MaNGA target has multiple viable companions galaxies, we chose the closest companion when measuring the pair's mass ratio and projected separation.

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. After restricting the sample to SFGs, we have 133 MaNGA targets with paired galaxies outside of the IFU. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample. This give use a total of 175 MaNGA targets in our pair sample.

@@

It has been shown that the SFR in galaxies is dependent on both the stellar mass and the redshift of the galaxies \citep{Noeske:2007}. We will compare our interacting galaxies with isolated galaxies of a similar stellar mass and redshift, to account for these other dependencies. To do this, we will use two different methods of pair - control comparison.

In the first method, we only control for the stellar mass between the pairs and controls. We split both the pair and control samples into five evenly spaced stellar mass bins over the range, \logm\ $=$ 9.0$-$11.5. The pairs are then compared to their respective controls within each stellar mass bin. While the redshift in not constrained in this method, we will show in Section \ref{sec:mass-bin} that this method is sufficient to reveal the merger induced star formation.

@@

For each MaNGA 1\arcsec$\times$1\arcsec\ spaxel, we use our spectral fitting code, {\sc spfit}, to get measurements of emission line fluxes, equivalent widths, velocities, and velocity dispersions of 19 emission lines and measurements of stellar masses, ages, [Fe/H], and the kinematics of the stellar populations.

We correct the emission lines for reddening using the extinction curve from \citet{Cardelli:1989} with updated coefficients from \citet{ODonnell:1994}. The extinction is parameterized as $R_V$ $\equiv$ $A_V/E(B-V)$ $=$ 3.1, where we estimate the value of the $V$-band extinction, $A_V$, by comparing the H$\alpha$/H$\beta$ ratio to the expected value of 2.85 for case-B recombination.

@@

We fit the geometry with apertures from both catalogs when available. When using the first method of pair-control comparison, where pairs and controls are grouped into stellar mass bins, we fit the control galaxies with the NSA apertures. When making the comparison with the second tailored control method, paired galaxies fitted with the NSA aperture are compared against controls fitted with the NSA aperture and paired galaxies fitted with the \citet{Simard:2011} apertures are compared against controls fitted with the \citet{Simard:2011} apertures. Further, when defining the a galaxy's geometry with the \citet{Simard:2011} apertures we also use the masses given in the catalog for the total stellar mass of the galaxy.

@@

@@

In the first method, paired and control galaxies are grouped into five evenly spaced stellar mass bins between \logm\ $=$ 9.0$-$11.5. Within each stellar mass bin, we create a median profile of log(sSFR) as a function of \reff-scaled galactocentric distance for both paired and control galaxies. The control profiles are constructed using all available control galaxies within each mass bin. The error associated with the ``stacked'' profile is the standard error of mean of the data at each radius bin. We show these stacked profiles for control galaxies in the left hand panel of Figure \ref{fig:ssfr_prof} and the stacked profiles for paired galaxies in the middle panel of Figure \ref{fig:ssfr_prof}. Star formation in the control galaxies are quenched in their centers with respect to their disks and the difference between the sSFR in their centers and their disks increases with stellar mass, consistent with previously published results using the MaNGA survey \citep{Belfiore:2018}. In contrast, the paired galaxies show flatter sSFR profiles where the level of the sSFR remains roughly consistent across their disks except for paired galaxies with stellar masses above \logm\ $=$ 10.5, which still show some central quenching.

@@

Alternatively, the central enhancement of the sSFR in the paired galaxies can be interpreted as the quenching of star formation in the centers of isolated galaxies. While the mechanism for quenching in galaxies remains under debate, the merger process seems to be capable of rejuvenating star formation in the centers of these galaxies, resulting in the flatter sSFR profiles.

@@

@@
